---
title: "Hypericum analysis"
author: "Marco Kreuzer"
output: html_document
---

## Description

This analysis is published in:
Howard C, Hill E, Kreuzer M, Mali P, Masiero E, Slater A, Sgamma T (2019) DNA Authentication of St John’s Wort (Hypericum perforatum L.) Commercial Products Targeting the ITS Region; Genes 10(4), 286

This file consists of the metabarcoding analysis after the querying the reads to
the NCBI database. The blast output was parsed with MEGAN. The starting data is the file
__specieslist.ncbi.txt__.

```{r, message=FALSE}
library(here)
library(tidyverse)
library(kableExtra)
```

## Analysis of NBCI BLAST output

```{r}
# Loading Data
hypericum_raw2 <- read.csv("specieslist.ncbi.txt", header = F)
names(hypericum_raw2) <- c("Sample", "Region","Read_name","Size","Taxon")

```

### Data Transformation

```{r}
# Select ITS1 and ITS2
hypericum_raw <-  hypericum_raw2 %>% select(Sample, Region, Size, Taxon, Read_name ) %>% 
  filter(Region == "ITS1" | Region == "ITS2") %>% ungroup()

# Add unique identifier
hypericum <- hypericum_raw %>% 
  mutate(Sample_id = paste(Sample, Region, sep = "_")) %>% 
  mutate(pkey = paste0(Sample_id,Taxon))

# calculate number of reads per sample
size_per_id <- hypericum %>% 
  select(Sample_id,Size) %>% 
  group_by(Sample_id) %>% 
  summarize(Total_size = sum(Size))

# add number of reads to each read and calculate its relative size
hypericum <- hypericum %>% 
  left_join(size_per_id, by="Sample_id") %>% 
  mutate(Rel_size = Size/Total_size) 
```

Filter Reads that are not assigned to a meaningful category or not assigned at all. Calculate nuber of reads matching to species in genus _Hypericum_.

```{r}
total_reads_not_assigned <- hypericum %>%   filter(Taxon == "cellular organisms" |
  Taxon == "environmental sample" |
  Taxon == "Eukaryota" |
  Taxon == "Pentapetalae" |
  Taxon == "Eudicotyledons" |
  Taxon == "Not assigned") %>% group_by(Taxon) %>% 
  summarise(Size = sum(Size)) %>% pull() %>% sum() 

# Total number of reads not assigned grouped by taxonomic unit
taxa_reads_notassigned <- hypericum %>% 
  filter(Taxon == "cellular organisms" |
  Taxon == "environmental sample" |
  Taxon == "Eukaryota" |
  Taxon == "Pentapetalae" |
  Taxon == "Eudicotyledons" |
  Taxon == "Not assigned") %>% 
  group_by(Taxon) %>% 
  summarise(Size = sum(Size)) %>% 
  mutate(percent = round(Size/total_reads_not_assigned*100, digits = 1))

# Total number of reads not assigned, grouped by Sample_id

taxa_reads_notassigned_sample <- hypericum %>% 
  filter(Taxon == "cellular organisms" |
  Taxon == "environmental sample" |
  Taxon == "Eukaryota" |
  Taxon == "Pentapetalae" |
  Taxon == "Eudicotyledons" |
  Taxon == "Not assigned") %>%
  group_by(Sample_id, Sample) %>% summarise(Size = sum(Size)) 

# number of Hypericum reads per sample
hypericum_reads <- hypericum %>% 
  filter(str_detect(Taxon, pattern = "Hyper")) %>% 
  group_by(Sample,Region) %>% 
  summarise(Size = sum(Size))

```

Summary stats table

```{r}
summary_stats <- hypericum %>% select(Sample, Region, Sample_id, Size, Rel_size) %>% 
  group_by(Sample_id, Sample, Region) %>% 
  summarise(Total_reads = sum(Size), Clusters = n(), Total_rel_size = sum(Rel_size)) #%>% 
# write.csv(x = summary_stats, file = "summary_stats_reads_clusters.csv")
kableExtra::kable(summary_stats)
```


### Data Preparation for Heat Map

```{r}
hypericum_hm <- hypericum %>% 
  group_by(Sample, Region, Taxon, Sample_id) %>% 
  summarise(Rel_size = sum(Rel_size), Clusters = n(), Size = sum(Size)) %>% 
  filter(Rel_size >= 0.02) %>% 
  ungroup() %>% 
  mutate(Sample = as.factor(Sample))

hypericum_hm <- hypericum_hm %>% filter(
  Taxon != "cellular organisms" &
  Taxon != "environmental sample" &
  Taxon != "Eukaryota" &
  Taxon != "Pentapetalae" &
  Taxon != "eudicotyledons" &
  Taxon != "Not_assigned" &
  Taxon != "Magnoliophyta"
)

hypericum_hm <- hypericum_hm %>% complete(Sample_id, Taxon)
# hypericum_hm %>% arrange(Taxon)

# Calculate Missing Data
missing_data <- hypericum_hm %>% select(Taxon, Clusters) %>% 
  group_by(Taxon) %>% 
  summarise(Sum = sum(Clusters, na.rm=T)) %>% 
  filter(Sum == 0) %>% select(Taxon) #%>% as.data.frame

hypericum_hm2 <- hypericum_hm %>% 
  anti_join(missing_data, by="Taxon") %>% 
  mutate(Sample = substr(Sample_id,1,3)) %>% 
  mutate(Region = substr(Sample_id, 5,8))

### Add a new categorie "combined" 
all_reads <- hypericum %>% select(Sample, Region, Sample_id, Size, Read_name) %>% 
  group_by(Sample) %>% 
  summarise(Total_reads = sum(Size), Clusters = n()) %>% 
  mutate(Sample = as.character(Sample))

hypericum_hm3 <- hypericum_hm2 %>% 
  mutate(Region = "combined") %>%
  left_join(all_reads, by = "Sample") %>% 
  select(Region, Taxon, Sample, Size, Total_reads) %>% 
  group_by(Region,Sample,Taxon, Total_reads) %>% 
  summarise(Size = sum(Size, na.rm = T), 
            Clusters = n()) %>%
  mutate(Rel_size = Size/Total_reads) %>% 
  mutate(Size = ifelse(Size == 0, NA, Size),
         Clusters = ifelse(Clusters == 0, NA, Clusters),
         Rel_size = ifelse(Rel_size < 0.02, NA, Rel_size))

hypericum_hm4 <- bind_rows(hypericum_hm2, hypericum_hm3)

```

Plot Heatmap

(note that the samples are renamed in the plots)

```{r}
hm <- ggplot(data = hypericum_hm4, aes(x = Region, 
                                       y = Taxon, 
                                       fill = Rel_size, 
                                       group = Sample)
             ) +
  geom_tile(stat = "identity", colour = "white") + 
  theme_bw() + 
  scale_fill_gradient(name = "Rel. Abund.", 
                      low = "antiquewhite1", 
                      high = "coral4", 
                      na.value = "azure1") +
  facet_grid(~Sample,
             labeller = labeller(Sample = c('826' = "218", 
                                            '827' = "224",
                                            '829' = "228",
                                            '830' = "229",
                                            '834' = "251")
                                 )
             ) +
  theme_grey(base_size=8) +
  theme(
    axis.text.y = element_text(face="italic"),
    axis.text.x = element_text(face = "bold", size = 6),
    axis.ticks=element_line(size=0.4),
    plot.background=element_blank(),
    panel.border=element_blank(),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    legend.title = element_text(size = 8),
    axis.title.x = element_blank(),
    strip.text.x = element_text(size = 8,
                                face = "bold")
    )

hm
```


```{r}
tiff(filename = "Heatmap.tiff", height = 4, width = 8, res = 300, units = "in")
hm
dev.off()
```

### Main taxonomic categories

How many reads mapped to Hypericum?

```{r}
hypericum_hm4 %>% group_by(Taxon, Sample_id, Sample, Region) %>% 
  summarise(Rel_size = round(sum(Rel_size, na.rm = T), digits = 2)) %>% 
  filter(str_detect(Taxon, 'Hyperic')) %>% 
  filter(Rel_size > 0) %>% 
  filter(Region == "combined")
```

# Local BLAST analysis

## Importing Local Blast Hits

Data Transformation

```{r}
blast.local <- read.table("samplelist_all_reanalysis.megablast")

names(blast.local) <- c("Sample","Region","qseqid", "Taxon", 
                        "pident", "length", "mismatch", 
                        "gapopen", "qstart", "qend", "sstart", 
                        "send","evalue","bitscore" )

blast.local <- blast.local %>% 
  mutate(Taxon = substr(Taxon, 7,100)) %>% 
  mutate(Sample = as.character(Sample)) %>% 
  mutate(Sample_id = paste(Sample,Region, sep = "_")) %>% 
  mutate(Size = str_extract(qseqid, "[^_size_]+$")) %>% 
  mutate(Size = str_remove(Size, ";")) %>%
  mutate(Size = as.numeric(Size)) %>% 
  mutate(qseqid = gsub("_size.*","",qseqid)) 

# Calculate total reads per cluster
total_reads <- blast.local %>% 
  select(Sample,Region, Size, qseqid) %>% 
  distinct() %>% 
  select(Sample,Region, Size) %>% 
  group_by(Sample, Region) %>% 
  summarise(Total_reads = sum(Size), count = n()) %>% 
  mutate(Sample_id = paste(Sample,Region, sep = "_")) %>% 
  ungroup() %>% 
  select(Sample_id, Total_reads)

# Add Rel_size
blast.local <- blast.local %>% left_join(total_reads, by = "Sample_id") 

blast.local <- blast.local %>% 
  mutate(Rel_size = Size/Total_reads)

blast.local %>% select(Rel_size, Sample_id) %>% 
  group_by(Sample_id) %>% summarise(Relative_size = sum(Rel_size))

blast.local %>% select(Rel_size, qseqid, Sample_id, Rel_size) %>% distinct() %>% 
  group_by(Sample_id) %>% 
  summarise(Rel_size = sum(Rel_size), count=n())
```

```{r}
pident_relabundance <- blast.local %>%
  filter(Region == "ITS1" | Region == "ITS2") %>% 
  group_by(Sample,Region, qseqid, Rel_size, Size) %>% 
  summarise(best_hit = max(pident))

# Hits from Hypericum NCBI

hypericum_ncbi <- hypericum %>% 
  filter(str_detect(string = Taxon, pattern = "Hyp")) %>% 
  select(Taxon, Read_name, Sample_id) %>% 
  mutate(NCBI = "Hit") %>% 
  rename(qseqid = Read_name) 

blast.local.ncbiInfo <- blast.local %>% 
  left_join(hypericum_ncbi, by="qseqid") %>% 
  mutate(NCBI = replace_na(NCBI, "No Hit"))

```

```{r}
pident_relabundance <- blast.local.ncbiInfo %>%
  filter(Region == "ITS1" | Region == "ITS2") %>% 
  group_by(Sample,Region, qseqid, Rel_size, Size, NCBI) %>% 
  summarise(best_hit = max(pident))

plot_pident_relabundance <- ggplot(data=pident_relabundance, aes(x = Rel_size, y = best_hit, colour = NCBI)) +
  geom_point() + 
  facet_grid(Region~.~Sample,
             labeller = labeller(
        Sample = c('826' = "218", 
                   '827' = "224",
                   '829' = "228",
                   '830' = "229",
                   '834' = "251"
                   ))) +
  ylab("Best Match per Cluster (% Sequence identity)") +
  xlab("Relative Abundance of Clusters") +
  scale_colour_manual(values = c("red", "darkgrey")) +
  theme(
    axis.title.y = element_text(),
    axis.title.x = element_text(),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    strip.text.x = element_text(face = "bold"),
    strip.text.y = element_text(face = "bold")
  )

plot_pident_relabundance

```


### Export Image
```{r}
tiff(filename = "Pident_Plot.tiff", height = 4, width = 6, res = 300, units = "in")
plot_pident_relabundance
dev.off()
```


```{r}
taxa_best_hits_pident <- blast.local.ncbiInfo %>% 
  group_by(Sample, Region, qseqid, NCBI) %>%
  filter(pident == max(pident)) %>% 
  arrange(Sample,Region,qseqid,Rel_size, bitscore, pident, Taxon.x) %>% 
  filter(NCBI =="Hit") %>% 
  select(Region, Sample, Taxon.x, Sample_id.x, bitscore, pident,qseqid, Rel_size)

write.csv(taxa_best_hits_pident, file="taxa_best_hits_pident99.csv")
getwd()

taxa_best_hist_rf <- read.csv("taxa_best_hits_NCBI.csv", header=T)
#taxa_best_hist_rf <- read.csv("taxa_best_hits_pident99.csv", header=T)

taxa_best_hist_rf <- taxa_best_hist_rf %>% 
  mutate(Sample = as.character(Sample))
taxa_best_hist_rf <- taxa_best_hist_rf %>% 
  filter(Region == "ITS1" | Region == "ITS2") %>% 
  filter(Rel_size > 0.02)
```


```{r}
hist_local_blast <- ggplot(data = taxa_best_hist_rf, 
                           aes(x=Region, y=Rel_size, fill=Taxon)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(name = "Taxa",
                    values = c("H_perforatum_type_1" = "chocolate3",
                               "H_perforatum_type_2" = "chocolate1",
                               "H_perforatum_type_3" = "brown1",
                               "H_perforatum_types123_H_attenuatum" = "burlywood4"),
                    labels = c("H_perforatum_type_1" = "H. perforatum Type 1",
                               "H_perforatum_type_2" = "H. perforatum Type 2",
                               "H_perforatum_type_3" = "H. perforatum Type 3",
                               "H_perforatum_types123_H_attenuatum" = "H. perforatum Types 1&2&3 \nand H. attenuatum")
                    ) +
  ylab("Cumulative Relative Abundance") +
  facet_grid(.~Sample,
             labeller = labeller(
        Sample = c('826' = "218", 
                   '827' = "224",
                   '829' = "228",
                   '830' = "229",
                   '834' = "251"
                   ))) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(
      size = 8,
      face = "bold",
      angle = 45,
      vjust = 1,
      hjust = 1),
    axis.text.y = element_text(
      size = 8,
      face = "bold"),
    legend.text = element_text(
      face = "italic"),
    strip.text.x = element_text(
      face = "bold")
  )

hist_local_blast

```

```{r}
tiff(filename = "Hypericum_histogram.tiff", height = 4, width = 6, res = 300, units = "in")
hist_local_blast
dev.off()
```
